Four decades of water quality change in the upper San Francisco Estuary\footnote{\Sexpr{vers}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.}

### 2017-02-28 sftrends_manu.Rnw
sort of done with jabusch edits

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{\Sexpr{vers}}}

@@

Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.  The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}. This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.  More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{BeckIP}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}. These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters. Application to additional datasets could provide further insight into environmental dynamics of aquatic systems.

@@

The \ac{sfe} drains a 200 thousand km$^2$ watershed and is the largest estuary on the Pacific coast of North America.  The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land from the Central Valley watershed.  Water enters \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate.  Water dynamics in \ac{sfe} are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.  Seasonally, inflows into \ac{sfe} peak in the spring and early summer from snowmelt in the upper watershed, whereas consumption, withdrawals, and export have steadily increased from 1960 to present but vary considerably depending on inter-annual climate effects \cite{Cloern12b}.  The system is mixed mesotidal and significant exchange with the ocean occurs daily, although the extent of landward saltwater intrusion varies with inflow and annual water use patterns. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.  Oceanic upwelling and climatic variation are also significant external factors that have influenced water quality dynamics in the Bay \cite{Cloern07}.

Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication (e.g., Chesapeake Bay).\cite{Kemp05}  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \cite{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \cite{Thompson08,Crauder16}.  However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \cite{Cloern07}, increased occurrence of hypoxic conditions \cite{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \cite{Lehman05,Lehman10}.  These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \cite{Cloern07}, biological invasions \cite{Cohen98}, and departures from the historical flow record \cite{Enright09,Cloern12b}.  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}

The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay (\cref{fig:delt_map}).  The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \cite{Jassby00,Jassby08,Novick15}.  Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged through the system with natural and anthropogenic processes.  A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}. For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.  Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.  Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \cite{Jassby00,Jassby08}.  Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

@@

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Stations were grouped by location in the study area for comparison: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \cite{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.  Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \cite{Beck15}.

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data. \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).  Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).  More sites had decreasing trends for the observed data from 1996-2013.  \Sexpr{cap(obsdir$din[2, 5])} sites had significant trends for \ac{din} (\Sexpr{obsdir$din[2, 3]} increasing, \Sexpr{obsdir$din[2, 4]} decreasing), \Sexpr{obsdir$nh[2, 5]} sites for ammonium (\Sexpr{obsdir$nh[2, 3]} increasing, \Sexpr{obsdir$nh[2, 4]} decreasing), and \Sexpr{obsdir$no23[2, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[2, 3]} increasing, \Sexpr{obsdir$no23[2, 4]} decreasing). Trends by location (upper Delta, middle, and Suisun stations) were not apparent, suggesting individual sites had trends that differed independent of relative location.  For example, P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites. Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).  Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects. As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed \ac{din} data at C10. Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b. In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.  The difference between predicted and flow-normalized results is shown in \cref{fig:dinc10}c, such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.  For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrients. Further, \cref{fig:dinc10}e shows \ac{wrtds} estimates of seasonal variation in the relationships of \ac{din} with flow throughout the period of record. Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.  Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s, which is potentially linked to long-term climatic patterns\cite{Cloern07}. 
<<echo = F, cache = F>>=
data(trnds_seasyr)

@@

# descriptions for magnitude changes
chgcmp <- select(cmp, -ztest) %>% 
  spread(dat, perchg) %>%
  mutate(
    signobs = sign(obs), 
    signnrm = sign(nrm),
    chgs = ifelse(
      signobs == -1 & nrm < obs, 'moreneg', 
      ifelse(signobs == 1 & nrm > obs, 'morepos',
      ifelse(signobs == 1 & signnrm == -1, 'postoneg', 
      ifelse(signobs == -1 & signnrm == 1, 'negtopos', 
      ifelse(signobs == -1 & signnrm == -1 & nrm > obs, 'lessneg', 
      ifelse(signobs == 1 & signnrm == 1 & nrm < obs, 'lesspos', 
        'somechg')
    )))))
  )
# table of results for shorter sexpr
chgtb <- table(chgcmp$chgs)

A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.  These comparisons are made to identify changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making. For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant. This suggests that time series that include flow effects have sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an increase in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positive or less negative) compared to observed trends.  \Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change and \Sexpr{eng(chgtb['somechg'])} sites went from no change to negative estimated trends for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).  The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.

>>>>

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{\Sexpr{vers}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.}}

@@

Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.  The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}. This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.  More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}. These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters. Application to additional datasets could provide further insight into environmental dynamics of aquatic systems.

@@

The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.  The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land from the Central Valley watershed.  Water enters \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).  Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.  Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \cite{Cloern12b}. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.

Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.  Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \cite{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \cite{Thompson08,Crauder16}.  However, recent water quality trends have suggested that resistence to nutrient inputs is decreasing given documented changes in chlorophyll biomass \cite{Cloern07}, increased occurrence of hypoxic conditions \cite{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \cite{Lehman05,Lehman10}.  These recent changes have been attributed to biological invasions \cite{Cohen98} and departures from the historical flow record \cite{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures).\cite{Cloern07}  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}

Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.  A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}. For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported.  Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.  Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \cite{Jassby00,Jassby08}.  Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

@@

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the northern \ac{SFE} region.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Stations were grouped by location in the study area for comparison: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \cite{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.  Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \cite{Beck15}.

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data. \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).  Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).  Decreasing trends were more common for the observed data from 1996-2013.  \Sexpr{cap(obsdir$din[2, 5])} sites had significant trends for \ac{din} (\Sexpr{obsdir$din[2, 3]} increasing, \Sexpr{obsdir$din[2, 4]} decreasing), \Sexpr{obsdir$nh[2, 5]} sites for ammonium (\Sexpr{obsdir$nh[2, 3]} increasing, \Sexpr{obsdir$nh[2, 4]} decreasing), and \Sexpr{obsdir$no23[2, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[2, 3]} increasing, \Sexpr{obsdir$no23[2, 4]} decreasing). P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section. Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).  Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.

<<echo = F, cache = F>>=
data(trnds_seasyr)

@@

# descriptions for magnitude changes
chgcmp <- select(cmp, -ztest) %>% 
  spread(dat, perchg) %>%
  mutate(
    signobs = sign(obs), 
    signnrm = sign(nrm),
    chgs = ifelse(
      signobs == -1 & nrm < obs, 'moreneg', 
      ifelse(signobs == 1 & nrm > obs, 'morepos',
      ifelse(signobs == 1 & signnrm == -1, 'postoneg', 
      ifelse(signobs == -1 & signnrm == 1, 'negtopos', 
      ifelse(signobs == -1 & signnrm == -1 & nrm > obs, 'lessneg', 
      ifelse(signobs == 1 & signnrm == 1 & nrm < obs, 'lesspos', 
      ifelse(signobs == 0 & signnrm == 1, 'zerotopos', 
      ifelse(signobs == 0 & signnrm == -1, 'zerotoneg',
      ifelse(signobs == -1 & signnrm == 0, 'negtozero', 
      ifelse(signobs == 1 & signnrm == 0, 'postozero', 
        NA)
    )))))))))
  )

# table of results for shorter sexpr
chgtb <- table(chgcmp$chgs)

@
Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects. As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends. For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant. This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an increase in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positive or less negative) compared to observed trends.  \Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change, \Sexpr{eng(chgtb['zerotoneg'])} sites went from no change to negative estimated change, and \Sexpr{eng(chgtb['zerotopos'])} site went from no change to a positive trend for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).  The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.
